---
title: "The effects of leaving a tip"
output: html_notebook
---

Do the practice exercize that is provided to us in question 9.6 of *Field Experiments*.

> Rind an Bordia studied the tiping behavior of lunchtime patrons of an "upscale Philadelphia restaurant" who were randomly assigned to  four experimental groups (Rand and Bordia 1996). One factor was server gender, and a second factor was whether the server draws a "happy face" on the back o the bill presented to customers. 

```{r setup, message=FALSE}
library(foreign)
library(data.table)
```

First, load the data. 

```{r}
d <- read.dta("http://hdl.handle.net/10079/vmcvdzs")
head(d)
```
There are only three fields that we care about in this table: 

- `female`: is the server a 'female' or 'male' server
- `happyface`: does the server, at random, place a happyface onto the bill of a patron
- `tip`: the amount of tip that is left on on the bill. 

# Questions to Answer

1. Write down a regression model that depicts the effect of the sex of the waitstaff, whether they put a happyface on the bill, and the interaction of these two terms. Test, using an `anova( ... , test = 'F')` for whether the interaction is necessary. This is going to compare a longer model to a shorter model. 
  1. How much does a female server make, on average? 
  2. How much does a male server make, on average? (Note, for both of these questions, think carefully about what the coefficients in your table are reporting. It requires more than a passing glance to interpret these correctly)
  


