---
title: "Sampling Distribution and Ranomization Inference"
output: github_document
---

```{r}
library(data.table)
library(ggplot2)

knitr::opts_chunk$set(dpi = 300)

set.seed(1)
```

# The sharp null through an Anodyne Example: Estrogen and Soybeans

Suppose that e are working through the example of a randomized
assignment of men and women to eat (or not eat) soybeans and we are
measuring the level of estrogen present in each of their bloodstreams
(perhaps in parts/million).

You *do not* have to work through this whole worksheet at once. In fact,
it might make more sense to only work through it alongside the examples
that David and David are talking about.

First, we create a grouping variable with two groups, one called “Man”,
and another called “Woman”.

```{r}
d <- data.table(id = 1:40)
d[ , group := c(rep("Man",20),rep("Woman",20))]
```

To these groups, we assign silly, but schematically helpful *potential
outcomes* to treatment and control. We say that, by some chance, we
sampled men into our study that had estrogen ppm levels that ranged from
1-20, in perfect increments. Also, what luck, but we sampled women into
the study that had estrogen levels that ranged from 51-70. So, the women
have on average higher estrogen beginning the study.

```{r}
d[ , po_control := c(seq(from = 1, to = 20), seq(from = 51, to = 70))]

# no effect because potential outcomes in treatment are the same
d[ , po_treatment := po_control + 0] 

head(d)
```

Per our randomization scheme, we are going to randomly assign the
individuals to either eat lots of tofu `(treatment == 1)` or to eat no
tofu `(treatment == 0)`. To do this, we write a simple function that
will randomly place zeros and ones for the treatment and control.

```{r}
randomize <- function(units_per_group) { 
  ## an (unnecessary) function to randomize units into 
  ## treatment and control 
  ## ---
  ## args: 
  ##  - units_per_group: how many zero and one should be returned
  
  assignment_vector <- rep(c('control', 'treatment'), each = units_per_group)
  sample(assignment_vector)
} 
```

```{r}
randomize(4)
```

```{r}
table(randomize(4))
```

With our randomization function in hand, we can now set up our vector of
treatment assignments. This is simply storing the results of our
function randomize in a vector object called treatment.

```{r}
d[ , condition := randomize(.N/2)]
```

Check that this has worked as intended.

```{r}
d[1:5, ]
```

```{r}
d[ , table(condition)]
```

Recall that we are setting up an experiment that has **no** effect. As
Green and Gerber point out in *Field Experiments* in the case of the
sharp-null, we are actually testing against the possibility that we
observe both the potential outcomes! (As a comprehension check, explain
why this is true).

Next, we create a vector of realized outcomes. first using the compact
notation that Green and Gerber prefer using maths. For those randomized
to treatment, we multiply the potential outcome to treatment time the
treatment vector (which is a 1 when they were assigned to treatment),
and for those in control, into this vector we assign the potential
outcome to control time the quantity `(1 - treatment)` which will be one
when they are in the treatment group.

```{r}
d[ , outcomes := (
  (I(condition == 'treatment') * po_treatment) + 
    (1 - I(condition == 'treatment') * po_control))
  ]
```

But, this is little different than an `ifelse` statement that produces
the same result.

```{r}
d[ , outcomes := ifelse(condition == 'treatment', po_treatment, po_control)]
```

Notice the nice concision that comes with these *vectorized* operations.
It is possible to write a process that produces the same result, but
that does so by looping through each row of the data.table to do so.

    outcomes <- rep(NA, length(group))
    for(i in 1:length(group)) { 
      if(treatment[i] == 'control') { 
        outcomes[i] <- po_control[i]
        } else if(treatment[i] == 'treatment') { 
          outcomes[i] <- po_treatment[i]
          }
      }

# Conducting an experiment is sampling (once) the treatment vector

Now that we have the data set up, we can begin to examine what the
lecture is really about, what is the distribution of ATE that we observe
due to the different possibly assignments to treatment and control.

A few points to remember:

1.  When we randomly assign people to a group – either treatment or
    control – we are producing an i.i.d. assignment process. **As a
    result,** for any concept that you might care about, and for any
    statistic calculated against that concept, this random assignment of
    people will produce at least a consistent estimate. *In the case of
    the sample mean, this will produce an unbiased estimate*.
2.  When we **intervene** in the system to provide people treatment or
    control we *reveal* and then *measure* either their potential
    outcome to treatment or their potential outcome to control.

This does not imply that any one realization of treatment/control
assignment and measurement will provide the exact value of the ATE.
Instead, it means that across all treatment and control assignment
vectors that are possible, the average of these vectors will be the true
ATE.

In this data, we can write an estimator that naturally follows the
two-step process described above.

First,

```{r}
group_averages <- d[ , .(group_mean = mean(outcomes)), keyby = .(condition)]
```

Then,

```{r}
ate <- group_averages[ , group_mean[condition == 'treatment'] - 
                         group_mean[condition == 'control'] ]
```

We can use the R base function `diff` to compute the differences in
means of the treatment and control groups. This `diff` function just
computes how much changes between positions in a vector.

```{r}
diff(c(1,2,4,7,11))
```

And so, with the `group_averages` object, we could call for,

```{r}
group_averages[ , diff(group_mean)]
```

Another useful feature of working in this `data.table` paradigm is that
we can aggregate data.tables just as we have done to produce the
`group_mean` variable. But, this aggregation is *itself* a data.table,
and so it is possible to continue to produce further aggregations –
namely to produce this difference we have just computed.

So, the entire process can be represented in a transparent one-line
call.

```{r}
d[ , .(group_mean = mean(outcomes)), keyby = .(condition)][ , diff(group_mean)]
```

**But wait\!** We created this data such that there is *exactly* zero
treatment effect. Note, even more specifically than creating the data so
that there was no *average treatment effect* we constructed this so that
there was no effect at all – this is at the potential outcomes level\!
Recall that $Y_{i}(0) \equiv Y_{i}(1), \forall i$. \