---
title: "w241: Experiments and Causality"
subtitle: "Unit 1"
author: "David Reiley, David Broockman, D. Alex Hughes"
institute: "UC Berkeley, School of Information"
date: "Updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(ggplot2)
library(data.table)

theme_set(theme_minimal())
```

Demo of math: 

$$
\begin{align*}
Y &= \beta_{0} + \beta_{1} x_{1} + \epsilon \\ 
  &= 2
\end{align*}
$$

---

# Importance of Experimentation 

- Large scale datasets are nice... but the majority of this data is not useful for assessing causal questions: _Does this cause that?_
- More data is not a solution for causal questions: 
  - More of the wrong type of data is not better than less of the wrong type of data. 
  - **Example**: randomized controlled trials ("RCTs")
    - **Treatment Group**: A group of subjects provided a new drug to determine efficacy -- but compared to whom? 
    - **Control Group**: A group of subjects provided _no_ new drug -- compared to the treatment group.
- Results from a well-constructed randomized experiment -- even if small -- contains information that even the largest, most dynamic observational data cannot. 

---

# Hormone Replacement Therapy (HRT)

## Reading Assignment

lease read [this article](https://www.nytimes.com/2007/09/16/magazine/16epidemiology-t.html) about _Hormone Replacement Therapy_ 
## Think about the following questions as you read

- What is the difference in _data generation_ strategy between the two studies? 
  - **Nurses Health Study** _key finding_: HRT reduces heart attacks among postmenopausal women
  - **Women's Health Initiative** _key finding_ HRT increases heart disease (as well as strokes and breast cancer). 
  
## The two results are not compatible

- Which of these results do be _believe_ and why? 
- HRT either increases, decreases, or doesn't change risk of heart attacks. How then can the Nurses' Health Study and Womens' Health Initiative disagree? 

---

# WHI results are more credible. Why? 

## Nurses' Health Study 

  - Epidemiological, (i.e. observational)
  - Large sample of women 
  - Statistically significant correlations 
  - "_Women who (choose to) take HRT are less likely to have heart attacks_."
  
## Women's Health Initiative 

  - Designed experiments 
  - Smaller sample size (vis-a-vis Nurses' Health Study)
  - "_Women who (are randomly assigned to) take HRT are more likely to have heart attacks._"
  
## Unmeasured confound

  - Nurses' Health Study: Those who _choose_ to take HRT have different underlying health conditions and motivations than those who _choose not_ to take HRT. 
  - Womens' Health Initiative: Those who _are assigned_ to undergo HRT are statistically indistinguishable form those who _are assigned_ **to not** undergo HRT. 
  
---

# The oat bran example

## Unmeasured confounders 

  - Potential invisible or unmeasurable factors that affect results 
  - These factors can be avoided by conducting the experiment with identical populations 
  
---

# Topic overview for the week 

  - Observation vs. intervention 
  - What experiments can tell us
  - Kinds of experiments in the "natural" and "social" sciences. 
  - _Example_: "Magic on the Internet" auction experiments 
  - _Reading Assignment_: Read _Field Experiments_ (Gerber and Green, 2012) Chapter 1.
  
---

# Reading: Key Points 

  - Causal questions are crucial in a varietey of areas 
    - Business 
    - Public policy 
    - Individual decisionmaking 
  - Decisions should be concerned with counterfactuals 
    - We observe the state of the world if we have done `X`. What would we observed of we had done `not X`? Or, if we had done `Y`? 
  - This causal inference is difficult because we cannot observe _both_ of these states of the world. 
  - Arguments based on intuition and anecdotes provide no solid ground for disputation and resolution. They usually end in stalemates. 
    - Should the government extend unemployment benefits if there is as pandemic that causes a recession? Perhaps this could be settled with an experiment. 
    - Should our organization continue to purchase display ads? 
    
---

# Reading: Key Points, cont'd 

  - Causal questions are settled with experiments in a way that avoid stalemates. 
  - Causal questions are harder to get correct in social and data sciences than in physical science, because of underlying heterogeneity 
    - All electrons are the same. 
    - Not all humans are the same. 
    - In fact, perhaps all humans are distinct. 
  - One should be skeptical of causal inferences based on observational (i.e. non-experimental) data because of the possibility of unobserved heterogeneity. 
  
---

# Examples of causal questions 

  - Does boiling drinking water prevent people from contracting cholera? 
  - Do mandatory seat belt laws reduce traffic fatalities? 
  - Does TV advertising increase purchases? If so, by how much? 
  - Does Chanel No. 5 have a downward-sloping demand curve? 
  - Does having children give people a more satisfying life? 
  
---

class: inverse, center, middle

# **Experiment**: An intervention that creates variation in order to teach us causal questions.

---

# Lemonade Stand Exmple, Part 1 

```{r make lemonade, echo = FALSE, message = FALSE, fig.align='center', dev='svglite', fig.height=4}
lemons <- data.table(id = 1:10)

lemons[ , price := runif(10, 0.25, 1.5)] %>% 
  .[ , quantity := 5 - 1.5 * price + rnorm(.N, sd = .15)]

lemons %>% 
  ggplot() + 
  aes(x = price, y = quantity) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + 
  labs(
    title = 'Lemonade Price and Quantity', 
    x = 'Price', y = 'Quantity'
  )
```

---

# Lemonade Stand Example, Part 2

```{r sour lemons, echo = FALSE, message = FALSE, fig.align='center', dev='svglite', fig.height=3}
sour_lemons <- data.table(id = 1:10)

sour_lemons[ , price := runif(10, 0.49, 0.51)] %>% 
  .[ , quantity := 5 - 1.5 * price + rnorm(.N, sd = .15)]

sour_lemons %>% 
  ggplot() + 
  aes(x = price, y = quantity) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + 
  coord_cartesian(xlim = c(0,1)) + 
  labs(
    title = 'Lemonade Price and Quantity', 
    x = 'Price', y = 'Quantity'
  )
```

  - We cannot learn anything about the demand curve for lemonade if the price remains the same! 
  - To learn about the demand curve, it is necessary for there to be _some_ variation in prices. 
  - When we conduct an experiment, we **deliberately create variation, but in a way that is not related to other features.**  
  
---

# Intervention and Randomization 

  - Experiments do _not_ have to involve randomization
  - Intervention is the key element of an experiment. 
  - Randomization can sometimes be difficult to implement
    - Randomly assign price to every customer? 
    - Might take too much time
    - Might irritate customers 
  - But we learn more -- and have better guarantees -- with deliberate variation than without it. 
  
---

# Why is randomization useful? 

## Consider the lemonade stand example

  - Charging a different price every day
  - Charging more on a very hod day, and more lemonade sold
    - Temperature is a confounding factor: like price, termperature also affects lemonade sales
    - Might incorrectly assume the price increase (rather than the temperature) caused the increase in sales. 
  - One wouldn't learn the true effects of a price change if any factor that also influences sales is correlated with price change. 
  
---

# Why is randomization useful? (cont'd) 

  - We can control for confounds by repeating the experiment many times 
  - The treatment, _price_, should be independent of every other feature that might influence lemonade sales: temperature, sun, holidays 

## Randomization produces the guarantee of independence. 

---

```{r experimental lemons, echo = FALSE, message = FALSE, fig.align='center', dev='svglite', fig.height=4}
lemons <- data.table(id = 1:10)

lemons[ , price := runif(10, 0.25, 1.5)] %>% 
  .[ , quantity := 1 + 1.5 * price + rnorm(.N, sd = .15)]

lemons %>% 
  ggplot() + 
  aes(x = price, y = quantity) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + 
  labs(
    title    = 'Lemonade Price and Quantity', 
    subtitle = 'Price Randomly Assigned', 
    x = 'Price', y = 'Quantity'
  )
```

---

# Pitfalls of naturally occurring variation

## Possible pitfalls in making causal inferences from observational data

  - Nurses' Health Study on HRT: 
    - _Possible story_: Women who receive hormones generally tended to care more about their health, and/or follow doctors' recommendations. 
    - Lower incedence of heart attacks possibly caused by _these_ factors, ratehr than HRT. 
  - Lemonade stand data
    - The higher the price, the more lemonade sold
    - Turns out, the kid is a ruthless capitalist! They charged more when the temperature was higher. 
    - Temperatuer is a confounding factor in this case. 
---

# What is a natural experiment? 

  - Green and Gerber are not fond of using the term, _natural experiment_. 
 - A _natural experiment_ is naturally occurring data -- i.e. the researcher did not themselves design the data -- that a researcher reasons has the same properties as a true, designed and controlled experiment. 
 
---
 
# An analogy: Herschel's Gardern 

- William Herschel: astronomer who discovered Uranus
- Experiments on stars are not possible -- but one can still learn something by studying various stars at different stages of development
- **Herschel's Garden**: the idea of viewing the night sky as a garden that features the same types of _plants_ at different stages of development
  - Perhaps this idea can be applied into the social data sciences
  - Natural experiments in the social data sciences are likely to be less clear, due to un-observable heterogeneity  
  
---

# Observational studies 

## Best 
- Data from _naturally occurring experiments_ (i.e. situations where variation was produced by something like random assignment) 
  - Example: charter school lottery deciding applicant acceptance
    - Data about applicants possibly analyzed to study causal effects of charte school education 
    - Same characteristics in both groups, those who attened the charger school and those who did not
  - Example: Vietnam draft lottery that causd some people to get more college education
    - Group that got more college education is otherwise identical to group that did not 
    
---

# Observational studies, cont'd

## Less ideal 
- Decent reasons to believe that those who received an intervention are otherwise identical to those who did not 
  - Example: Want to know the effect of plowing snow on business in a college town. Can't randomly assign plowing, but the plows clear different streets on alternating days. The days of the week might not be otherwise identical in terms of foot traffic. 
  - Example: Does raising the minimum wage _actually_ cause a decrease in employment? We cannot randomly set the minimum wage, but Seattle raised its minimum wage to $15, while Tacoma did not. 
  
---

# Observational studies, cont'd 

## Non-credible 

- No reason to believe that those who received an intervention are otherwise identical to those who did not 
  - Example: Charging more for lemonade on hot days than cool days; women receiving HRT care more about their health. 
  
---
class: inverse, center, middle

# Magic on the internet

---

# Magic on the internet 

- Using field experiments to test equivalence between auction formats 
- Observational data holds no guarantees that auction format is independent of card value. 

---

# Early Online Magic Auctions 

- People trading _Magic_ cards via Usenet newsgroup
- New transaction mechanism, before eBay
  - Currency: Cash 
  - Question: How should one value a _Magic_ card? 
- Beginning of _Magic_ auctions, with bids updated via email 
- Cards sold -- auction closed -- if there were no price increasese in three days. 

---

# Why conduct this experiment? 

- Auction _Magic_ cards in a controlled experiment to learn about _auction theory_. 
  - Note that the context is magic cards, but the larger question is broader than this particular good; rather it is about auctions as a concept. 
- Why is auction theory important? 
- Why study revenue equivalence between auction formats? 
- Which format generates higher selling prices? Ascending-bid format, or sealed-bit format? 
- There isn't actually a prediction about which should generate higher revenues, in expectation. 

---

# Lab researchc on auctions 

- Students bid on a fictitious good -- a good that they don't particularly want, and are not particularly invested in
- So experimenters, assign different values to different subjects 
  - Each subject knows their own value, but they do not know the values that others assign to the good 
  - Goal is to maximize the difference between the price you pay, and the value that you ascribe to the good. 
  - Importantly, there is _no_ value in this system that is beyond the assigned value. Does this match reality? 
- With this setup, researchers can examine ascending-bid and sealed-bid auctions 
- Find interesting violations of theory 
- These lab violations raised the question: are these inconsistencies with theoretical predictions an artifact of the labratory, or are these more general when people are acting within a _real_ auction market? 

---

# Vickrey: Auction Formats 

1. English auctions: ascending bids 
2. Sealed-bid auctions: one-time highest bids
3. Dutch auctions: decending price clock 
4. Vickrey's "second-price" auction: second-highest bid determines winner's price (like eBay proxy bidding)

---

# Revenue Equivalence: Two Kinds

- Strategic equivalence: strong prediction
  - Dutch and first-price auctions are strategically equivalent because they have the same amount of information 
  - English and second-price auctions are strategically equivalent under _private value_ (like those assigned in the lab)
    - Thus, there is a dominant-strategy mechanism for trutheful revelation of valuations. 
    - Regardless of others' strategies, one's own optimal stretegy is the same bid: bid my maximum willingness to pay 
- General revenue equivalence: weaker prediction 
  - Expected revenue of all four formats should be the ame if people are risk-neutral
  
---

# Lab research: Violation to theory 

## Cox et al. (1982, 1983)

- First price auctions raise more revenue than Dutch uactions 

## Kagel et al (1987, 1993) 

- With private values, subjects overbid in second-price auctions 
- So, yielded higher revenues than English auctions 

## Why? 

- Could this be risk aversion? 

---

# Field Experiments 

## Similar to observational studies 

- Auctions for real good, by real people, who are really accustomed to bidding for these goods 

## Similar to lab experiments 

- Deliberate, designed interventions into a _real_ system, without actors being aware of the interventions. 
- Two different auction formats with the same good and the same bidders. 

## Similar to real world 

- Cannot control individuals valuations of the good 
- Cannot control risk, time, or other preferences of the subjects

---

# Background on Magic: The Gathering 

- First sold in July 1993, with a first printing of 10 million cards 
- To date, more than 20 billion cards printed 
- Estimated 1995 wholesale revenues: greater than $100 million USD 
- Cards are sold in a random assortment, which generates a large aftermarket 
- Creates a real world market laboratory 

---

# The experiment 

- Four pairs of auctions designed for within-card comparisons 
- Auctioned sold the same card twice (e.g. in both a Dutch and first-price auctions
- Sealed-bid auction: one week to submit bids 
- English auction: 
  - Bid any time 
  - Daily update on each card's high bid 
  - Like the Usenet marketplace, three days without a new bid closes the auction

---

# The experiment, cont'd 

- Dutch auction: start at a high price, announce a decrease of 5% each day via e-mail; same bid increments as other auction formats 
- To control for order effects -- perhaps there is only a market for one card, and so once it sells, the market has little demand for another -- each experiment was run twice 
  - FD: First-price followed by Dutch 
  - DF: Dutch followed by first-price 
  - FD and DF sets were run at the same time 

---

# Result 1: Card-level data 

- Violates Dutch/first-price revenue equivalence 
- FD and FD experiments 173 matched pairs of cards 
  - Dutch format: 122 yielded higher revenue 
  - First-price format: 34 yielded higher revenue 
- On average, cards yielded $0.32 more in Dutch auction (24% of total card value) 
- Sign-rank test: price-per-card differences are highly significant 
- No qualitative difference between FD and DF results; that is, order of treatment doesn't matter 
- Opposite of the violation that had been observed in the lab. 

---

# Result 2: Bid-level data 

- Treatment unit: what is an observation? 
- Here: an individual bid, vs. a card the receives multiple bids
- Bid-level data weakly supports (note the loose language here) violation of strategic equivalence: Dutch and first price auctions should have generated the same bids; but some evidence this might not be true. 
- Compare bids by the same bidder in two matched auctions 
- Data censoring -- we don't see most people's revealed prices because there is only one bid that is made, which closes the auction
- Of 38 observations with bids observed in both Dutch and first-price auctions: 
  - 30 favored the Dutch; mean difference $2.52
  - 4 favored first-price, mean difference -$0.50
  
---

# Result 3: English auction 

- Seems to produce slightly more revenue than second-price auction 
- Card level data: 164 pairs observed 
  - 1.8% more revenue than second-price auction (not statistically significant difference) 
  - Cannot reject null hypothesis of revenue equivalence 
- Bid-level data: 112 matched bid pairs 
  - 75 cases had higher bids in English auction 
  - 29 cases had higher bids in second-price 
  - On average, English bids were 3% higher 
- Estimates opposite to lab results, but with limited statistical power 

---

# Result 4: First/Dutch vs. English/Second 

- See how auction revenues deviate from reference price ("cloister price") for each good 
- Pool together Dutch/first-price data and English/second price data: total of 370 observations 
- On average DF auctions raise 12% more revenue than ES auctions
- Lower difference for higher-priced cards; cards above $13 show no differences 
- Results are consistent with prior lab research 
- Could this be risk aversion? If so, then why is the effect smaller for higher-priced cards? 

---

# Conclusions 

## Revenue ranking 

1. Dutch 
2. First-price 
3. English 
4. Second-price 

## Field data vs. laboratory data 

- Opposite violations of the FD and ES strategic violations from the lab results 
- Same FD > ES effects as lab results 

---

# Questions 

- What cause results tobe different from those in the lab? 
  - Real goods mechanism? 
  - Cash payout mechanism? 
  - Simultaneous vs. sequential mechanism? 
  - Clock speed mechanism? 
- New lab experiments (Katok and Kwasnica, 2003): Slower clock speeds lead to higher revenue in Dutch auctions -- people are impatient? 

---
